\name{}: machine learning over interaction networks and ontology-based background knowledge predicts drug--target interactions

### 2021-04-21 paper/main.tex
Submit revisioned paper

% With the rapidly growing field of machine learning approaches and
% their application to bioscientifical problems in the realm of
% bioinformatics, different kinds of data, such as long DNA sequences
% could be utilized for feature generation, while rapid advances were
% made. 
Identifying drug--target interactions (DTIs) is important in drug
repurposing and repositioning, and in the exploration of novel drug
candidates and their interaction partners. Computational methods are
widely applied to predict DTIs and many computational methods have
been developed.  These methods can be broadly classified into
``top-down'' and ``bottom-up'' approaches. Top-down approaches start
form observable characteristics resulting from a drug--target
interaction, such as side-effects or the diseases treated by a drug,
and infer likely molecular mechanisms (i.e., the interaction) using
these observations.
% and indications represented by knowledge graphs or ontologies, induced by a drug and infer targets based on the likely molecular mechanisms that result in these phenotypes
Bottom-up approaches start from molecular features such as molecular
structure or fingerprints associated with drug and protein, and
predict interactions from this information.

Both bottom-up and top-down approaches to drug--target interaction
prediction have some advantages and limitations. Generally, bottom-up
methods face the challenge to predict whether a structure binds to a
protein given their molecular properties; whether two entities
interact depends not only on the molecular structure of the entities
(where binding sites and molecular forces need to be determined for
accurate prediction) but also properties such as where a protein is
expressed or in which celltypes and anatomical structures. Top-down
methods use information for DTI prediction such as side-effect
similarity \cite{Campillos2008} that is largely complementary to
knowledge gained from molecular properties. While methods that rely on
molecular information are directly predicting whether two molecular
could interact, top-down methods rely on more indirect means and 
infer a DTI from observable effects resulting from the interaction.

Both approaches may be used in conjunction with network inference
\cite{Chen2015}. Biological networks used for DTI prediction include
protein--protein interaction networks \cite{Feng2017, Lee2018} and
networks including several other types of biological relations,
including similarity between represented entities \cite{Ding2013,
  Gottlieb2011}. Network-based DTI prediction methods use the
guilt-by-association principle \cite{Oliver2000} and assume that a
protein is a likely target for a drug if many of the protein's
neighbors in the interaction network are targets of the drug
\cite{Gillis2012}. Network-based methods have been applied
successfully to DTI prediction. However, if DTIs are taken as direct
physical interactions between a drug and protein, it remains an
unresolved question whether the network-based guilt-by-association
hypothesis is true, or whether an interaction of a drug and protein
dys-regulates several of the protein's interaction partners, and
therefore resulting in effects that are not direct interactions but
only downstream consequences of an interaction.

Progress in machine learning using graph neural networks can allow us
to test this hypothesis and combine both bottom-up and top-down
features with a network in a single machine learning model. 
In particular, Graph Convolutional Networks \cite{GCNConv} and
their variants using different types of kernels \cite{ChebConv,
  ARMAConv}, including attention mechanisms \cite{GATConv}, and
different forms of exploring node neighborhoods \cite{APPNPConv,
  SAGEConv} can combine different types of features and graph-based
information. % While based on diverse systems,
% they can be relevant for testing distinct hypothesis for given
% graphs. While convolutional filters are suitable for finding patterns
% among the given graph, attention mechanisms are more relevant for
% discovery of important regions within. Lately, graph learning
% approaches found application for computing compound representations
% for DTI prediction.
They have previously been applied for a number of tasks, including
prediction of protein functions \cite{Zitnik2017}, cancer drug
response \cite{Liu2020} and drug--target affinity prediction
\cite{GraphDTA2020}.

One challenge in DTI prediction are potential biases resulting from
the underlying datasets \citep{Pahikkala2014} which may affect model
evaluation and comparison. First, novel drugs are often designed by
altering non-functional components of a drug, leading to two and more
very similar drugs designed to target the same proteins
\cite{Overington2006}. This can result in a bias when it leads to
hidden duplicates or highly similar compounds that are distributed
among training and evaluation dataset, resulting in a better
(measured) predictive performance than would be expected when the
model is applied to identify drugs that target a protein for which no
drugs yet exist. Second, some proteins (which we call \textit{hub
  proteins}) have significantly more known interactions with drugs
than others. In the STITCH database, $5\%$ of the proteins have $40\%$
of the interactions, and similar distributions are present in other
datasets \cite{Drugbank2007, Drugbank2017}; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (i.e.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed). This may
affect common evaluation schemes\cite{Survey2018} where it is possible
to exploit these biases within DTI prediction \cite{}. \citet{vanLaarhoven2014} showed that another bias can be heavily capitalized over the dataset of \citet{Yamanishi2008}.

We developed \name{} as a method for predicting DTIs. We use an
ontology-based machine learning method \cite{DL2vec2020} to encode
phenotypic consequences of DTIs and deep learning methods to encode
molecular features. We combine both using a protein interaction
network which we exploit using a graph neural network. We use this
model to test whether molecular or phenotype features benefit from the
network information and find that only phenotype features localize on
the graph whereas molecular features do not. We further evaluate and
compare \name{} against several DTI prediction methods and demonstrate
a substantial improvement of \name{} over the state of the art in
predicting drugs that target a protein. We also identify and
characterize several biases in both training and evaluating DTI
prediction methods, and make recommendations on how to avoid
them. \name{} is available as Free Software at
\url{https://github.com/THinnerichs/DTI-VOODOO}.

@@

As phenotypes and functions are encoded through ontologies, we use
DL2Vec \citep{DL2vec2020} to obtain ontology based representations for
use as top-down features. DL2vec constructs a graph by introducing
nodes for each ontology class and edges for ontology axioms, followed
by random walks starting from each node in the graph. These walks are
encoded using a Word2vec \citep{Word2vec2013} model. Therefore, DL2Vec
generates representations that can encode drug effects or protein
functions while preserving their semantic neighbourhood within that
graph.

@@

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with a
contrastive loss using cosine similarity.  A half-twin neural network
aims to learn a similarity between two embeddings of the same
dimension.  The precomputed embeddings may have different
dimensionality and we first process them using a fully connected
neural network layer which takes as input an embedding and outputs a
representation of a particular size, i.e., we use this layer as a
trainable feature transformation and apply it to reduce the
representation size of the embeddings for drugs and proteins
separately.  An example structure for both types of features can be
found in Figure~\ref{fig:HalfTwinNetwork}.
% While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we
The use of this trainable feature transformation layer enables
flexible experimentation as both ontology and molecular feature for
both drugs and proteins are reduced to the same dimensionality for
varying sizes of inputs; this allows for a high amount of modularity
across different experimental setups by adding different kinds of
features into the model. Additionally, the generated features may be
used for other tasks. % ; for
% example, the ontology LFT can be reused for a variety of DL2vec-based
% features with respect to other ontologies and hypotheses
We follow the results of \textit{DL2vec} \cite{DL2vec2020} and use as
activation function $\sigma := \mathrm{LeakyReLU}$ which leads to
improved performance compared to other activation functions.

@@

While in this initial formulation of a GCNConv the node-wise update
step is defined by the sum over all neighbouring node representations,
we can alter this formulation to other message passing schemes.  We
can rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$, and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We denote this
generalized layer type as \textsc{GENConv} following the notation of
PyTorch Geometric \cite{PytorchGeometric}.  While the reordering is
mainly import for numerical stability, this alteration also addresses
the vanishing gradient problem for deeper convolutional networks
\cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale and highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously. % , and the resulting information\todo{Which information?
  % Specify} hold no information on their own connectivity.
We further employ another mechanism to avoid redundancy and fading
signals in stacked graph convolutional networks, using residual
connections and a normalization scheme \cite{DeepGCN2019,
  DeeperGCN2020}.  The residual blocks are reusable and can be stacked
multiple times. % The structure of the GNN architecture we use is shown
% in Figure~\ref{fig:ResGraphConvBlocks}.

@@

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparse with respect to the number of
both drugs and proteins considered, the training, validation and
testing datasets are imbalanced. As there are only $22,336$ links in
the considered STITCH subset, the ratio
\begin{equation}
	w:= \frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
	\label{weight}
\end{equation}
consequently needs suitable compensation in the computed loss function and
appropriate metrics for the evaluation.

@@

We construct a graph with proteins as nodes and protein-protein
interactions as edges, mapping the protein features to each target as
node features. \name{} then propagates information among the PPI network
utilizing graph convolutional steps, calculates the similarity of drug
and protein representations, and predicts whether there is an
interaction. The full workflow scheme is depicted in
figure~\ref{fig:ModelWorkflow}

@@

Another view on these two approaches is as direct and indirect ways to
predict drug--target interactions. Molecular information can be used
to directly determine whether two molecules (such as a drug and
protein) have the ability to interact, whereas information about
phenotypic consequences of a drug (drug effects) or disruption of a
protein function can be used to indirectly suggest candidate
drug--target interactions. Molecular information will be specific to a
drug--target pair and we would not expect this information to
propagate through a protein--protein interaction network; the main
information about drug--target interactions that could be obtained
from interactions between proteins is information about binding sites
between proteins that may also be used by a drug molecule (i.e.,
information that protein $P_1$ binds to protein $P_2$ reveals
information about the molecular structures of both $P_1$ and
$P_2$). On the other hand, phenotypic consequences of changes in
protein function, or drug effects, are often a result of aberrant
pathway or network activity and involve more than one protein;
consequently, we expect these features to benefit more from including
information about protein--protein interactions. Our results (Table
\ref{tab:Results}) confirm this hypothesis and demonstrate that
molecular features do not benefit from including the interaction
network whereas the indirect, top-down features benefit from the
propagation over the interaction network.

@@

There are other types of indirect features that could be added to our
model. A common feature that may be added are drug indications which
are predictive of drug--target interactions \cite{Gottlieb2011};
however, we do not include them in our model as including drug
indications would allow our model to make many trivial predictions
based only on remembering which targets are often used for which
indication; including network information would likely benefit
predictions based on drug indications because different drugs may
target the same pathway through different mechanisms.

>>>>

% With the rapidly growing field of machine learning approaches and
% their application to bioscientifical problems in the realm of
% bioinformatics, different kinds of data, such as long DNA sequences
% could be utilized for feature generation, while rapid advances were
% made. 
Identifying drug--target interactions (DTIs) is important in drug
repurposing and repositioning, and in the exploration of novel drug
candidates and their interaction partners. Computational methods are
widely applied to predict DTIs and many computational methods have
been developed.  These methods can be broadly classified into
``top-down'' and ``bottom-up'' approaches. Top-down approaches start
from observable characteristics resulting from a drug--target
interaction, such as side-effects or the diseases treated by a drug,
and infer likely molecular mechanisms (i.e., the interaction) using
these observations.
% and indications represented by knowledge graphs or ontologies, induced by a drug and infer targets based on the likely molecular mechanisms that result in these phenotypes
Bottom-up approaches start from molecular features such as molecular
structure or fingerprints associated with drug and protein, and
predict interactions from this information.

Both bottom-up and top-down approaches to drug--target interaction
prediction bear some advantages and limitations. Generally, bottom-up
methods face the challenge to predict whether a structure binds to a
protein given their molecular properties; whether two entities
interact depends not only on the molecular structure of the entities
(where binding sites and molecular forces need to be determined for
accurate prediction) but also properties namely where a protein is
expressed or in which celltypes and anatomical structures. Top-down
methods use information for DTI prediction such as side-effect
similarity \cite{Campillos2008} that is largely complementary to
knowledge gained from molecular properties. While methods that rely on
molecular information are directly predicting whether two molecular
could interact, top-down methods base on more indirect means and 
infer a DTI from observable effects resulting from the interaction.

Both approaches may be used in conjunction with network inference
\cite{Chen2015}. Biological networks used for DTI prediction include
protein--protein interaction networks \cite{Feng2017, Lee2018} and
networks including several other types of biological relations,
including similarity between represented entities \cite{Ding2013,
  Gottlieb2011}. Network-based DTI prediction methods use the
guilt-by-association principle \cite{Oliver2000} and assume that a
protein is a likely target for a drug if many of the protein's
neighbors in the interaction network are targets of the drug
\cite{Gillis2012}. Network-based methods have been applied
successfully to DTI prediction. However, if DTIs are taken as direct
physical interactions between a drug and protein, it remains an
unresolved question whether the network-based guilt-by-association
hypothesis is true, or whether an interaction of a drug and protein
dysregulates several of the protein's interaction partners, and
therefore resulting in effects that are not direct interactions but
only downstream consequences of an interaction.

Progress in machine learning using graph neural networks can allow us
to test this hypothesis and combine both bottom-up and top-down
features with a network in a single machine learning model. 
In particular, Graph Convolutional Networks \cite{GCNConv} and
their variants operate on different types of kernels \cite{ChebConv,
  ARMAConv}, including attention mechanisms \cite{GATConv}, and
different forms of exploring node neighborhoods \cite{APPNPConv,
  SAGEConv} can combine different types of features and graph-based
information. % While based on diverse systems,
% they can be relevant for testing distinct hypothesis for given
% graphs. While convolutional filters are suitable for finding patterns
% among the given graph, attention mechanisms are more relevant for
% discovery of important regions within. Lately, graph learning
% approaches found application for computing compound representations
% for DTI prediction.
They have previously been applied for a number of tasks, including
prediction of protein functions \cite{Zitnik2017}, cancer drug
response \cite{Liu2020} and drug--target affinity prediction
\cite{GraphDTA2020}.

Potential biases resulting from
the underlying datasets \citep{Pahikkala2014} which may affect model
evaluation and comparison pose a challenge for DTI prediction. Firstly, novel drugs are often developed by
altering non-functional components of a drug, leading to two and more
very similar drugs designed to target the same proteins
\cite{Overington2006}. This can result in a bias when it leads to
hidden duplicates or highly similar compounds that are distributed
among training and evaluation dataset, resulting in a better
(measured) predictive performance than it would be expected when the
model is applied to identify drugs that target a protein for which no
drugs yet exist. Secondly, some proteins (which we call \textit{hub
  proteins}) have significantly more known interactions with drugs
than others. In the STITCH database, $5\%$ of the proteins have $40\%$
of the interactions, and similar distributions are present in other
datasets \cite{Drugbank2007, Drugbank2017}; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (i.e.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed). This might
affect common evaluation schemes\cite{Survey2018} where it is possible
to exploit these biases within DTI prediction \cite{}. \citet{vanLaarhoven2014} showed that another bias can be heavily capitalized over the dataset of \citet{Yamanishi2008}.

We developed \name{} as a method for predicting DTIs. We use an
ontology-based machine learning method \cite{DL2vec2020} to encode
phenotypic consequences of DTIs and deep learning methods to encode
molecular features. We combine both using a protein interaction
network which we exploit with the aid of a graph neural network. We use this
model to test whether molecular or phenotype features benefit from the
network information and find that only phenotype features localize on
the graph whereas molecular features do not. We further evaluate and
compare \name{} against several DTI prediction methods and demonstrate
a substantial improvement of \name{} over the state of the art in
predicting drugs that target a protein. We also identify and
characterize several biases in both training and evaluating DTI
prediction methods, and make recommendations on how to avoid
them. \name{} is available as Free Software at
\url{https://github.com/THinnerichs/DTI-VOODOO}.

@@

As phenotypes and functions are encoded through ontologies, we use
DL2Vec \citep{DL2vec2020} to obtain ontology based representations for
use as top-down features. DL2vec constructs a graph by introducing
nodes for each ontology class and edges for ontology axioms, followed
by random walks starting from each node in the graph. These walks are
encoded using a Word2vec \citep{Word2vec2013} model. Therefore, DL2Vec
generates representations that enable to encode drug effects or protein
functions while preserving their semantic neighborhood within that
graph.

@@

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with a
contrastive loss using cosine similarity.  A half-twin neural network
aims to learn a similarity between two embeddings of variable but same
dimension. As the original feature space may have varying dimensionality, we first process them using a fully connected
neural network layer which takes as input an embedding and outputs a
representation of a particular size, i.e., we use this layer as a
trainable feature transformation and apply it to reduce the
representation size of the embeddings for drugs and proteins
separately.  An example structure for both types of features can be
found in Figure~\ref{fig:HalfTwinNetwork}.
% While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we
The use of this trainable feature transformation layer enables
flexible experimentation as both ontology and molecular feature for
both drugs and proteins are reduced to the same dimensionality for
varying sizes of inputs; this allows for a high amount of modularity
across different experimental setups by adding different kinds of
features into the model. Additionally, the generated features may be
used for other tasks. % ; for
% example, the ontology LFT can be reused for a variety of DL2vec-based
% features with respect to other ontologies and hypotheses
We follow the results of \textit{DL2vec} \cite{DL2vec2020} and use as
activation function $\sigma := \mathrm{LeakyReLU}$ which leads to
improved performance compared to other activation functions.

@@

While in this initial formulation of a GCNConv the node-wise update
step is defined by the sum over all neighboring node representations,
we can alter this formulation to other message passing schemes.  We
can rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$, and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We denote this
generalized layer type as \textsc{GENConv} following the notation of
PyTorch Geometric \cite{PytorchGeometric}.  While the reordering is
mainly important for numerical stability, this alteration also addresses
the vanishing gradient problem for deeper convolutional networks
\cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale and highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously. % , and the resulting information\todo{Which information?
  % Specify} hold no information on their own connectivity.
We further employ another mechanism to avoid redundancy and fading
signals in stacked graph convolutional networks, using residual
connections and a normalization scheme \cite{DeepGCN2019,
  DeeperGCN2020}.  The residual blocks are reusable and can be stacked
multiple times. % The structure of the GNN architecture we use is shown
% in Figure~\ref{fig:ResGraphConvBlocks}.

@@

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparse with respect to the amount of
both drugs and proteins considered, the training, validation and
testing datasets are imbalanced. As there are only $22,336$ links in
the considered STITCH subset, the ratio
\begin{equation}
	w:= \frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
	\label{weight}
\end{equation}
consequently needs suitable compensation in the computed loss function and
appropriate metrics for the evaluation.

@@

We construct a graph with proteins as nodes and protein-protein
interactions as edges, mapping the protein features to each target as
node features. \name{} then propagates information among the PPI network
utilizing graph convolutional steps, calculates the similarity of drug
and protein representations, and predicts whether there is an
interaction. The full workflow scheme is depicted in
Figure~\ref{fig:ModelWorkflow}

@@

Another view on these two approaches is as direct and indirect ways to
predict drug--target interactions. On one hand, molecular information can be used
to directly determine whether two molecules (such as a drug and
protein) have the ability to interact, whereas information about
phenotypic consequences of a drug (drug effects) or disruption of a
protein function can be used to indirectly suggest candidate
drug--target interactions. Molecular features will be specific to a
drug--target pair and we would not expect this information to
propagate through a protein--protein interaction network; the main
information about drug--target interactions that could be obtained
from interactions between proteins is information about binding sites
between proteins that may also be used by a drug molecule (i.e.,
information that protein $P_1$ binds to protein $P_2$ reveals
information about the molecular structures of both $P_1$ and
$P_2$). On the other hand, phenotypic consequences of changes in
protein function, or drug effects, are often a result of aberrant
pathway or network activity and involve more than one protein;
consequently, we expect these features to benefit more from including
information about protein--protein interactions. Our results (Table
\ref{tab:Results}) confirm this hypothesis and demonstrate that
molecular features do not benefit from including the interaction
network whereas the indirect, top-down features benefit from the
propagation over the interaction network.

@@

There are other types of indirect features that could be added to our
model. A common feature that may be added are drug indications which
are predictive of drug--target interactions \cite{Gottlieb2011}.
However, we do not include them in our model as including drug
indications would allow our model to make many trivial predictions
based only on remembering which targets are often used for which
indication; including network information would likely benefit
predictions based on drug indications because different drugs may
target the same pathway through different mechanisms.
